A Generalized Logical Framework

### 2025-03-11 yonedaLF/types2025/abstract.tex
types25 abstract typo

Logical frameworks (LFs \cite{Harper93lf}) and the closely related two-level type
theories (2LTTs \cite{twolevel}) let us work in a mixed syntax of a metatheory and a
chosen object theory. Here, we have a second-order view on the object theory,
where contexts, variables and substitutions are implicit, and binders are
represented as meta-level functions. There are some well-known limitations to
LFs. First, we have to pick a model of the object theory externally. Second,
since we only have a second-order view on that model, many constructions cannot
be expressed; for example, the induction principle for the syntax of an object
theory requires a notion of first-order model, where contexts and substitutions
are explicit. Various ways have been described to make logical frameworks more
expressive by extending them with modalities (e.g.\ \cite{sterlingthesis,DBLP:conf/lics/Hofmann99,DBLP:journals/corr/abs-1901-03378,orton_et_al:LIPIcs:2016:6564}). In the current work we
describe an LF with the following features:
\begin{itemize}
\item We can work with multiple models of multiple object theories at the same
  time. By ``theory'' we mean a second-order generalized algebraic theory
  (SOGAT \cite{uemura,DBLP:conf/fscd/KaposiX24}); this includes all type theories and programming languages that
  only use structural binders.
\item We have both an ``external'' first-order view and an ``internal'' second-order view on each model,
      and we can freely switch between perspectives.
\item Models of object theories are themselves internally defined in the LF.
\item The LF is fully structural as a type theory; no substructural modalities
  are used.
\end{itemize}

@@

  $\PSh\,i$. Then, we have $\In : \{i : \Base\} \to \Cat\,i \to \MetaTy$ and $\msf{base} : \In\,C \to \Base$.

@@

  any $C : \Cat\,j$, i.e.\ starting from a category that's internal to any

@@

    & \Y : \El\,(\Sub_M\,\Gamma\,\Delta) \simeq (\{j : \In\,M\} \to \El\,(\Y\,\Gamma) \to \El\,(\Y\,\Delta))\\

@@

\noindent \textbf{Sketch of the semantics}. The model of LF is constructed in
two steps. First, we give a model for the theory that has $\PSh$, $\Base$ and
$\In$ as sorts but does not support $\MetaTy$, and then take presheaves over
that model to obtain a model of a 2LTT where $\MetaTy$ represents the outer
layer. In the inner model, we start with an inductive definition of certain
``trees in categories'':
\begin{alignat*}{3}
  & \mbf{data}\,\Tree\,(B : \Cat) : \msf{Set}\,\mbf{where}\\
  & \hspace{1em} \msf{node} : (\Gamma : \PSh\,B)(n : \mbb{N})(C : \Fin\,n \to \Fib\,(B \ext \Disc\,\Gamma))\\
  & \hspace{3.8em} ((i : \Fin\,n) \to \Tree\,(B \ext \Disc\,\Gamma \ext C\,i)) \to \Tree\,B
\end{alignat*}
Here, $\PSh$ means presheaves in sets, $\Fib$ is cartesian fibrations, $\Disc$
creates a discrete fibration from a presheaf and $\blank\!\ext\!\blank$ takes
the total category of a fibration. Now, the objects of the semantic base
category are elements of $\Tree\,1$, and morphisms between trees are level-wise
natural transformations between the $\Gamma$ components together with $\Fin\,n
\to \Fin\,m$ renamings of subtree indices. A semantic $\Base$ points to a
subtree of a context, an $\In$ is a $\Fin\,n$ index pointing to a child of a
given node, and a $\PSh$ is a dependent presheaf a $\Gamma$ inside a given node.
This semantics can be viewed as a way of inductively carving out a subcategory
of $\mbf{Cat}$ where ETT type formers in presheaf categories are strictly stable
under substitution.

>>>>

Logical frameworks (LFs \cite{Harper93lf}) and the closely related two-level type
theories (2LTTs \cite{twolevel}) let us work in a mixed syntax of a metatheory and a
chosen object theory. Here, we have a second-order view on the object theory,
where contexts, variables and substitutions are implicit, and binders are
represented as meta-level functions. There are some well-known limitations to
LFs. First, we have to pick a model of the object theory externally. Second,
since we only have a second-order view on that model, many constructions cannot
be expressed; for example, the induction principle for the syntax of an object
theory requires a notion of first-order model, where contexts and substitutions
are explicit. Various ways have been described to make logical frameworks more
expressive by extending them with modalities (e.g.\ \cite{sterlingthesis,DBLP:conf/lics/Hofmann99,DBLP:journals/corr/abs-1901-03378,orton_et_al:LIPIcs:2016:6564}). In the current work we
describe an LF with the following features:
\begin{itemize}
\item We can work with multiple models of multiple object theories at the same
  time. By ``theory'' we mean a second-order generalized algebraic theory
  (SOGAT \cite{uemura,DBLP:conf/fscd/KaposiX24}); this includes all type theories and programming languages that
  only use structural binders.
\item We have both an ``external'' first-order view and an ``internal''
  second-order view on each model, and we can freely switch between
  perspectives. All models of object theories are defined internally in the LF.
\item The LF is fully structural as a type theory; no substructural modalities
  are used.
\end{itemize}

@@

  $\PSh\,i$. Then, we have $\In : \{i : \Base\} \to \El\,(\Cat\,i) \to \MetaTy$ and $\msf{base} : \In\,C \to \Base$.

@@

  any $C : \El\,(\Cat\,j)$, i.e.\ starting from a category that's internal to any

@@

    & \Y : \El\,(\Sub_M\,\Gamma\,\Delta) \simeq (\{j : \In\,M\} \to \El\,(\Y\,\Gamma\,\{j\}) \to \El\,(\Y\,\Delta\,\{j\}))\\

@@

\noindent \textbf{Sketch of the semantics}. The model of LF is constructed in
two steps. First, we give a model for the theory that has $\PSh$, $\Base$ and
$\In$ as sorts but does not support $\MetaTy$, and then take presheaves over
that model to obtain a model of a 2LTT where $\MetaTy$ represents the outer
layer. In the inner model, we start with an inductive definition of certain
``trees in categories'':
\begin{alignat*}{3}
  & \mbf{data}\,\Tree\,(B : \Cat) : \msf{Set}\,\mbf{where}\\
  & \hspace{1em} \msf{node} : (\Gamma : \PSh\,B)(n : \mbb{N})(C : \Fin\,n \to \Fib\,(B \ext \Disc\,\Gamma))\\
  & \hspace{3.8em} ((i : \Fin\,n) \to \Tree\,(B \ext \Disc\,\Gamma \ext C\,i)) \to \Tree\,B
\end{alignat*}
Here, $\PSh$ means presheaves in sets, $\Fib$ is cartesian fibrations, $\Disc$
creates a discrete fibration from a presheaf and $\blank\!\ext\!\blank$ takes
the total category of a fibration. Now, the objects of
the semantic base category are elements of $\Tree\,1$, and morphisms between
trees are level-wise natural transformations between the $\Gamma$ components
together with $\Fin\,n \to \Fin\,m$ renamings of subtree indices. The
non-discrete $\Fib$ components are preserved by morphisms. A semantic $\Base$
points to a subtree of a context, an $\In$ is a $\Fin\,n$ index pointing to a
child of a given node, and a $\PSh$ is a dependent presheaf over a $\Gamma$ inside a
given node. Extending a context with an $\In$ binding adds a new empty subtree
to a given node. Extending with an $\El$ binding extends the $\Gamma$ presheaf
in a node with a dependent presheaf.
